preregistration

### 2022-03-18 preregistrations/2_study1/preregistration.Rmd
preregistration for study 1

In line with previous theorizing [e.g., @ellis_2020; @frankenhuis_2016; @frankenhuis_2013], we initially expected people from adversity to perform better on cognitive tasks that require a broad attention scope in order to successfully detect peripheral and subtly changing information, and to perform worse on tasks requiring ignoring interfering information (see the [preregistration](https://github.com/StefanVermeent/attention_project/blob/Preregistration/preregistrations/1_pilot/preregistration.docx) of the first pilot study). However, the data pointed in the opposite direction. A Drift Diffusion analysis showed that participants with more violence exposure were generally slower to process information and orient their attention across all three tasks. At the same time, they were faster to narrow down their attention to the central target on the Flanker Task, thereby effectively limiting the amount of flanker interference that they were exposed to.

The findings in the pilot study suggest that deficit patterns that are typically found on tasks like the Flanker might not reflect problems in attention control and inhibition, even suggesting the opposite. Instead, people from adversity might process information more slowly, leading to longer reaction times. The goal of the current study is to replicate the attention findings on the Flanker Task, and to better understand the apparent information processing deficits. Participants completed three versions of the Flanker Task: the standard version, a version with enhanced visual information, and a version with degraded visual information.

@@

# Secondary  {#secondary}

@@

The Flanker Task was programmed in JsPsych version 3.6.1 [@deLeeuw_2015]. We programmed three conditions of the Flanker Task (See Figure \@ref(fig:figure1). In the standard condition, the arrows were 40 pixels in size (0.4 inches) and had zero padding between them. In the 'enhanced' condition, the aim was to increase the quality of visual information provided by the arrows. To this end, we increased the arrow size by 12.5% to 45 pixels (0.45 inches), and increased padding between the arrows to 5 pixels. This increased the width of the Flanker display by 50% with respect to the standard display. In the 'degraded' condition, the aim was to decrease the quality of visual information provided by the arrows. The arrow size and padding were the same as in the standard version, but both flanking and target arrows were rotated 45$^\circ$. Therefore, they provided more noisy information about their horizontal direction. We had no strong *a priori* expectations about how performance of adversity-exposed people might be affected by specific types of stimulus degradation (e.g., arrow rotation, decreasing stimulus contrast).

@@

In addition, we measured the stability of the family and social environment. On a scale of 1 (the same all the time) to 5 (constant and rapid changes), participants indicated how often the following aspects of their family and social environment changed before age 13: 1) economic status, 2) family environment, 3) childhood neighborhood environment, and 4) childhood school environment. Finally, we included four objective measures of unpredictability before age 13. Participants provided answers to the following items in an open response format: 1) "How often did you move?"; 2) "How many adults lived in your home on average?"; 3) "Think about the caregiver who had the most romantic partners before you were age 13. How many romantic partners did this person have?".

We computed four main unpredictability variables to be able to assess how they would relate to the dependent variables as well as to each other: 1) *unpredictability~perceived~*: containing all items of the perceived unpredictability scale and the QUIC, given their close conceptual overlap (*M* = `r #TBD`, *SD* = `r #TBD`, $\alpha$ = `r #TBD` *r* = `r #TBD` between the two scales); 2) *unpredictability~CHAOS~*: containing the items of the CHAOS (*M* = `r #TBD`, *SD* = = `r #TBD`, $\alpha$ = `r #TBD`); 3) *unpredictability~subjective~*: containing all items of the perceived unpredictability scale, the QUIC and the CHAOS (*M* = `r #TBD`, *SD* = `r #TBD`, $\alpha$ = `r #TBD`); 4) *unpredictability~objective*: containing the items assessing the stability of the family- and social environment , as well as the three objective measures (*M* = `r #TBD`, *SD* = `r #TBD`, $\alpha$ = `r #TBD`).

>>>>

In line with previous thinking (e.g., Ellis et al., 2020; Frankenhuis et al., 2016; Frankenhuis & Weerth, 2013), we initially expected people from adversity to perform better on cognitive tasks that require a broad attention scope. A broad attention scope might facilitate vigilance and the detection of subtle changes or peripheral information. However, this might come at a cost, especially for tasks that involve interference. To test these ideas, we ran a pilot study with a Cued Attention, Changed Detection, and Flanker Task (interference) and measured adversity exposure. We expected adversity-exposure would improve performance on the Cued-Attention and Change Detection tasks but hurt performance on the Flanker Task (see the preregistration of the first pilot study).

In contrast to expectations, our pilot data did not find evidence for a broad adversity-related attention scope. In fact, our Drift Diffusion analyses showed that participants with more violence exposure were generally slower to process information – as shown by the drift rate parameters of the Cued Attention and Change Detection Task and the perceptual input parameter of the Flanker Task - and orient their attention – As shown by the non-decision time parameter - across all three tasks. However, on the Flanker Task — where we expected lowered performance — our analyses revealed that people exposed to violence were faster to narrow their attention to the task-relevant information. In other words, instead of maintaining a broad scope of attention, people exposed to violence were able to quickly focus their attention, as shown by the attention shrinking parameter.

These findings are interesting for two reasons. First, the Flanker is a widely used task and people exposed to adversity typically show lowered performance. However, using the DDM to break performance into components, our pilot data suggest that lowered performance may not be caused by problems with attention control or inhibition. Instead, people exposed to adversity might simply extract perceptual information from the stimuli more slowly (as indicated by the perceptual input (p) parameter). This effectively lowers the quality of information that people can use to guide their decision, which will be reflected in lower reaction times. Second, these initial findings suggest that performance might be improved through different types of interventions focused on different cognitive mechanisms. Specifically, people with more adversity exposure might benefit from manipulations that increase the visual quality of stimuli.

The goal of the current study is to replicate the attention findings on the Flanker Task, and to better understand the apparent information processing deficits. Participants completed three versions of the Flanker Task: the standard version, a version with enhanced visual information, and a version with degraded visual information.

@@

# Secondary aims  {#secondary}

@@

The Flanker Task was programmed in JsPsych version 6.3.1 [@deLeeuw_2015]. We programmed three conditions of the Flanker Task (See Figure \@ref(fig:figure1). In the standard condition, the arrows were 40 pixels in size (0.4 inches) and had zero padding between them. In the 'enhanced' condition, the aim was to increase the quality of visual information provided by the arrows. To this end, we increased the arrow size by 12.5% to 45 pixels (0.45 inches), and increased padding between the arrows to 5 pixels. This increased the width of the Flanker display by 50% with respect to the standard display. In the 'degraded' condition, the aim was to decrease the quality of visual information provided by the arrows. The arrow size and padding were the same as in the standard version, but both flanking and target arrows were rotated 45$^\circ$. Therefore, they provided more noisy information about their horizontal direction. We had no strong *a priori* expectations about how performance of adversity-exposed people might be affected by specific types of stimulus degradation (e.g., arrow rotation, decreasing stimulus contrast).

@@

In addition, we measured the stability of the family and social environment. On a scale of 1 (the same all the time) to 5 (constant and rapid changes), participants indicated how often the following aspects of their family and social environment changed before age 13: 1) economic status, 2) family environment, 3) childhood neighborhood environment, and 4) childhood school environment. Finally, we included several objective proxy measures of unpredictability before age 13: 1) "How often did you move", 2) "Before age 13, did any other adults besides your caregivers live in your primary residence?". If the participant answered "yes" to this question, two additional questions asked about the number of different adult women (men) that lived in their primary residence for a period of time besides their primary caregiver. 3) If the participant indicated more adults living in the household besides their primary caregiver, this was followed up by items asking how many of these individuals were romantically involved with their caregiver.

four objective measures of unpredictability before age 13. Participants provided answers to the following items in an open response format: 1) "How often did you move?"; 2) "How many adults lived in your home on average?"; 3) "Think about the caregiver who had the most romantic partners before you were age 13. How many romantic partners did this person have?".

We computed four main unpredictability variables to be able to assess how they would relate to the dependent variables as well as to each other: 1) *unpredictability~perceived~*: containing all items of the perceived unpredictability scale and the QUIC, given their close conceptual overlap (*M* = `r #TBD`, *SD* = `r #TBD`, $\alpha$ = `r #TBD` *r* = `r #TBD` between the two scales); 2) *unpredictability~CHAOS~*: containing the items of the CHAOS (*M* = `r #TBD`, *SD* = = `r #TBD`, $\alpha$ = `r #TBD`); 3) *unpredictability~subjective~*: containing all items of the perceived unpredictability scale, the QUIC and the CHAOS (*M* = `r #TBD`, *SD* = `r #TBD`, $\alpha$ = `r #TBD`); 4) *unpredictability~objective*: containing the items assessing the stability of the family- and social environment, as well as the three objective measures (*M* = `r #TBD`, *SD* = `r #TBD`, $\alpha$ = `r #TBD`).
